---
title: "Simulating effects"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dagitty)
library(data.table)
library(knitr)
library(boot)
library(ggplot2)
```

## The data generating model


When we trying to estimate effects, it makes snese to think about the data generating model. Differently than simply assumeing an effect size for specific statistical tests, this implies formulating how we think the data were are planning to collect come about. This means more specifically to think about 

* observed (and unobserved variables)
* relationships between variables
* the strenght of these relationships

Here is my current understaning of the situation:

Trough a _TSST_ manipulation we can induce _negative affect_. A drug manipulation with an _opioid_ leads to a _reduction of negative affect_, which in turn leads to more _wanting_ but not nessesarily _liking_ (of opioid).

A visual representation of this model looks as follows:

```{r cars}
DGP = dagitty(
"dag {
  TSST    [pos=\"0,0\"]
  NEGAFF  [pos=\"1,0\"]
  OPIOID  [pos=\"1.5,1\"]
  REDNA   [pos=\"2,0\"]
  WANTING [pos=\"3,1\"]
  LIKING  [pos=\"3,-1\"]
  TSST -> NEGAFF
  NEGAFF -> REDNA
  OPIOID -> REDNA
  REDNA -> WANTING
  REDNA -> LIKING
}")
exposures(DGP) = c("TSST","OPIOID")
outcomes(DGP) = c("REDNA","WANTING","LIKING")
plot(DGP)
```

## Simulating data

In the simplest case, one could simply assume that all variables in the model are normally distributed. This assumptions does not seem to fit well, because the TSST and OPIOD manipulations are binary. Moreover, it is not clear if negatice affect is just the opposite pole of positive affect. This is important, becuase if negative affect were just one end of a one-dimensional construct, this would imply that if opioids reduce negative affect they should also increase positive affect. 

For now, we'll work under the assumption that opioids primarily reduce negative affect, without also strengthening positiuve affect. This means that we need to model `NEFAGG` wit a lower bound of zero.

Now lets just walk from right to left through the model. We start with two groups, one which does get the TSST intervention and one which does not get it.

```{r}
N = 100
TSST = sort(rep(c(0,1),N/2))
```

Next, we model negative affect as an effect of TSST plus som noise:
```{r}
NEGAFF = inv.logit(TSST*3+rnorm(N, mean = -2))
h = hist(NEGAFF,plot = F)
hist(NEGAFF[TSST == 1],
     breaks = h$breaks,
     xlim = range(h$breaks),
     border = NA,
     col = adjustcolor("red",alpha = .5),
     main = "Negative affect")
hist(NEGAFF[TSST == 0],
     breaks = h$breaks,
     xlim = range(h$breaks),
     border = NA,
     col = adjustcolor("blue",alpha = .5),
     add = T)
```

The `ìnv.logit` function is the same inverse logit that is also used in logistic regressions. we use this function to get from uncontrained normally distributed values to ratings of negative affect that are bound between zero and a maximum value. 


Next we define the opioid intervention, such that participants with and without TSST intervention receive opioids

```{r}
OPIOID = rep(sort(rep(c(0,1),
                      N/4)),
             2)
image(rbind(TSST,OPIOID),
      xaxt = "n",
      yaxt = "n")
axis(3, at = c(0,1),
     labels = c("TSST","OPIOID"),
     lwd = 0,
     cex = 2)
```


Now we can define reduction in TSST as an effect of initial negative affect plus effect of the opioid intervention (plus some random error):
```{r}
REDNA = NEGAFF - inv.logit(logit(NEGAFF) - OPIOID + rnorm(N))
```

And we plot to see if we get the expected results:

```{r}
dt = data.table(TSST,OPIOID,NEGAFF,REDNA)
dt[,OPIOID := factor(OPIOID)]
dt[,TSST := factor(TSST)]
tbl = dt[,list(mREDNA = mean(REDNA), mNEGAFF = mean(NEGAFF)), by = c("TSST","OPIOID")]
kable(tbl, digits = 2)

ggplot(tbl,aes(x = OPIOID, y = mREDNA, group = TSST, color = TSST)) + geom_line()
```


Now we can repeat this process 1000 times to see a distribution of effects:

```{r}
K = 1000
OPIOID = rep(sort(rep(c(0,1),
                      N/4)),
             2)

TSST = sort(rep(c(0,1),N/2))
reduction = matrix(NA,ncol = 2, nrow = K)
colnames(reduction) = c("no TSST","TSST")
for (k in 1:K) {
  NEGAFF = inv.logit(TSST*3+rnorm(N, mean = -2))
  REDNA = NEGAFF - inv.logit(logit(NEGAFF) - OPIOID + rnorm(N))
  dt = data.table(TSST,OPIOID,NEGAFF,REDNA)
  dt[,OPIOID := factor(OPIOID)]
  dt[,TSST := factor(TSST)]
  tbl = dt[,list(mREDNA = mean(REDNA), mNEGAFF = mean(NEGAFF)), by = c("TSST","OPIOID")]
  
  reduction[k,] = tbl[,diff(mREDNA), by = "TSST"][["V1"]]
}

delta = reduction[,"TSST"]-reduction[,"no TSST"]
hist(delta, 
     main = paste("% correct sign of effect:",round(mean(delta > 0),digits = 2)))
abline(v = 0, col = "red")

```


So we see that with the current paramters we are not even sure that the sign of the effect (larger effect of opioids after TSST) is is correct.
There are several things we can do. 

* discuss if the DGP reflects what you think
* gather more information about the effect size of the TSST and opioid manipulation

Once this is done, we can extend the simulations to include uncertainty about effect sizes and downstream effects, like the effect on wanting vs. liking.

